---
title: "Continuous Assignment 1"
author: 
  - "Fhatuwani Mokwenda"
  - "Jadene Jaftha"
format:
  pdf: 
    mainfont: "Times New Roman"
    documentclass: article
    geometry: a4paper, top = 2cm, bottom = 2cm, left = 2cm, right = 2cm

fontsize: 11pt

engine: knitr
---

## Question 1

```{r}
#| echo: false
#| message: false
#| warning: false

# for libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra) 
library(broom)
library(glmnet)
library(caret)
library(pROC)
library(xgboost)

```

```{r}
#| echo: false
#| message: false
#| warning: false
#Split data: train_dfing (80%) and testing(20%)
set.seed(2026)
listings <- read.csv("listings.csv")
train_dfindex = sample(nrow(listings), size = 0.8*nrow(listings))
train_df = listings[train_dfindex, ]
test = listings[-train_dfindex, ]
```

Exploring the *Price* variable, we see that it is heavily left-skewed. Because of this, we choose a high percentile value in order for the cutoff to represent the most "balanced" variation.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| out-width: "0.9\\textwidth"

# picking a cuttoff x
Price <- train_df$price
cutoff <- 0.75

x <- quantile(Price, cutoff)
hist(Price, 
     main = "Distribution of Price",
     col = "lightblue",
     breaks = 50)
abline(v = x, col = "red")

```

```{r}
#| echo: false
#| message: false
#| warning: false

train_df$HighPrice = as.integer(train_df$price > x)

#Get rid of original variable
train_df <- train_df |> select(-price)

test <- test |>  select(-price)

#Proportion Table 
prop.table(table(train_df$HighPrice)) 

```

```{r}
#| echo: false
#| message: false
#| warning: false

#Defining variables 
y = train_df[, 30]
X = train_df[, c("host_listings_count", "latitude", "longitude", "accommodates", "beds", "review_scores_rating")]
train_df_pros = data.frame(X, HighPrice = y)
lm_full = lm(HighPrice ~ ., data = train_df_pros)
#summary(lm_full)
```
This analysis is performed by looking at *HighPrice* as the categorical response variable where 
$$
HighPrice =
\begin{cases}
1 & \text{if price exceed the } 75^{th} \text{ percentile} \\
0 & \text{otherwise}
\end{cases}
$$
The predictor variables are as follows:
...
\newpage

## Question 2
### Modelling Approach
```{r}
#| echo: false
#| message: false
#| warning: false
# centering for KNN and elastic net
prepoc <- preProcess(X, method = c("center", "scale"))
X_processed <- predict(prepoc, X)
train_proc <- cbind(X_processed, HighPrice = y) #rebuilding training set
train_proc$HighPrice <- factor(train_proc$HighPrice,
                               levels = c(0, 1),
                               labels = c("No", "Yes")) #changing HighPrice to Yes or No

# Cross Validation (definition)
train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

modelLookup("rf")
```



#### Regularised Elastic-Net Logistic Regression
```{r}
#| echo: false
#| message: false
#| warning: false
alpha <- seq(0, 1, 0.1) # 0: ridge and 1: lasso
lambda_vals <- 10 ^ seq(-4, 1, length = 20)
enet_fit <- train(
  HighPrice ~ .,
  data = train_proc,
  method = "glmnet",
  trControl = train_control,
  tuneGrid =expand.grid(alpha = alpha, lambda = lambda_vals),
  metric = "ROC"
)
enet_fit

```

#### Random Forest
```{r}
#| echo: false
#| message: false
#| warning: false

set.seed(2026)
rf_fit <- train(
  HighPrice ~ .,
  data = train_proc,
  method = "ranger",
  tuneLength = 3,
  num.trees = 150,
  importance = 'impurity',
  trControl = train_control,
  num.threads = parallel::detectCores()
)
rf_fit

```

#### xgBoost
```{r}
#| echo: false
#| message: false
#| warning: false

set.seed(2026)
xb_grid <- expand.grid(
  nrounds = c(50, 100, 150),
  max_depth = c(3, 5, 7),
  eta = c(0.01, 0.1, 0.3),
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

xb_fit <- train(
  HighPrice ~ .,
  data = train_proc,
  method = "xgbTree",
  trControl = train_control,
  tuneGrid = xb_grid,
  metric = "ROC"
)

```


#### KNN
```{r}
#| echo: false
#| message: false
#| warning: false


# tuning
knn_fit <- train(
  HighPrice ~ .,
  data = train_proc,
  method = "knn",
  tuneGrid = expand.grid(k = 1:20),
  trControl = train_control,
  metric = "ROC"
)
```
\newpage

### Model Performance

Using a threshold of $\tau = 0.5$, we're comparing each model's metrics for **Accuracy**, **F1 Score**, **Precision**, **Recall**, **Specificity**, and **AUROC**.

```{r}
#| echo: false
#| message: false
#| warning: false
models <- list(ElasticNet = enet_fit, RF = rf_fit, XGBoost = xb_fit, KNN = knn_fit)

metrics <- function (model){
  pred <- model$pred
  
  # select best-tuning params
  if (!is.null(model$bestTune)){
    for (col in names(model$bestTune)){
      pred <- pred[pred[[col]] == model$bestTune[[col]], ]
    }
  }
  
  pred$pred_class <- factor(ifelse(pred$Yes >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
  
  # Confusion Matrix
  cm <- confusionMatrix(pred$pred_class, pred$obs, positive = "Yes")
  
  # AUROC
  roc_obj <- roc(pred$obs, pred$Yes)
  # return
  c(
    Accuracy = round(cm$overall["Accuracy"], 3),
    F1 = round(cm$byClass["F1"], 3),
    Precision = round(cm$byClass["Precision"], 3),
    Recall = round(cm$byClass["Recall"], 3),
    Specificity = round(cm$byClass["Specificity"], 3),
    AUROC = round(auc(roc_obj), 3)
  )
}

results <- sapply(models, metrics) |> t() |> as.data.frame()

# Rename duplicated columns
colnames(results) <- c("Accuracy","F1 Score","Precision","Recall","Specificity","AUROC")

# Add model names
results$model <- rownames(results)
results <- results[, c("Accuracy","F1 Score","Precision","Recall","Specificity","AUROC")]
results
```

```{r fig.width=12, fig.height=10,fig.pos='H'}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| out-width: "0.9\\textwidth"

## plotting
long_results <- results |>
  tibble::rownames_to_column(var = "Model") |> 
  pivot_longer(
    cols = -Model,
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(long_results, aes(x = Metric, y = Value, colour = Model, group = Model)) +
  geom_point(size = 3) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(
    title = "Comparison of Model Performance Metrics.",
  ) +
  scale_color_brewer(palette = "Set2")

```


### Summary
which model performs the best.

\newpage

## Question 3


\newpage

## Question 4
