---
title: "Continuous Assignment 1"
author: 
  - "Fhatuwani Mokwenda"
  - "Jadene Jafthar"
format:
  pdf: 
    mainfont: "Times New Roman"

engine: knitr
---

## Question 1

```{r}
#| echo: false
#| message: false
#| warning: false

# for libraries
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra) 
library(broom)
library(glmnet)
library(caret)

```

```{r}
#| echo: false
#| message: false
#| warning: false
#Split data: train_dfing (80%) and testing(20%)
set.seed(2026)
listings <- read.csv("listings.csv")
train_dfindex = sample(nrow(listings), size = 0.8*nrow(listings))
train_df = listings[train_dfindex, ]
test = listings[-train_dfindex, ]
```

Exploring the *Price* variable, we see that it is heavily left-skewed. Because of this, we choose a high percentile value in order for the cutoff to represent the most "balanced" variation.

```{r}
#| echo: false
#| message: false
#| warning: false

# picking a cuttoff x
Price <- train_df$price
cutoff <- 0.75

x <- quantile(Price, cutoff)
hist(Price, 
     main = "Distribution of Price",
     col = "lightblue",
     breaks = 50)
abline(v = x, col = "red")

```

```{r}
#| echo: false
#| message: false
#| warning: false

train_df$HighPrice = as.integer(train_df$price > x)

#Get rid of original variable
train_df <- train_df |> select(-price)

test <- test |>  select(-price)

#Proportion Table 
prop.table(table(train_df$HighPrice)) 

```

```{r}
#| echo: false
#| message: false
#| warning: false

#Defining variables 
y = train_df[, 30]
X = train_df[, c("host_listings_count", "latitude", "longitude", "accommodates", "beds", "review_scores_rating")]
train_df_pros = data.frame(X, HighPrice = y)
lm_full = lm(HighPrice ~ ., data = train_df_pros)
#summary(lm_full)
```
This analysis is performed by looking at *HighPrice* as the categorical response variable where 
$$
HighPrice =
\begin{cases}
1 & \text{if price exceed the } 75^{th} \text{ percentile} \\
0 & \text{otherwise}
\end{cases}
$$
The predictor variables are as follows:
...
\newpage

## Question 2
```{r}
#| echo: false
#| message: false
#| warning: false
# centering for KNN and elastic net
prepoc <- preProcess(X, method = c("center", "scale"))
X_processed <- predict(prepoc, X)
train_proc <- cbind(X_processed, HighPrice = y)
```



### Regularised Elastic-Net Logistic Regression
```{r}
#| echo: false
#| message: false
#| warning: false
alpha <- seq(0, 1, 0.1) # 0: ridge and 1: lasso
lambda_vals <- 10 ^ seq(-4, 1, length = 20)
enet_fit <- train(
  HighPrice ~ .,
  data = train
)

```

### Random Forest
```{r}
#| echo: false
#| message: false
#| warning: false
```

### xgBoost
```{r}
#| echo: false
#| message: false
#| warning: false
```


### KNN
```{r}
#| echo: false
#| message: false
#| warning: false

# Cross Validation
train_control <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

# tuning
knn_fit <- train(
  HighPrice ~ .,
  data = train_proc,
  method = "knn",
  tuneGrid = expand.grid(k = 1:20),
  trControl = train_control,
  metric = "ROC"
)
  
```

\newpage

## Question 3

\newpage

## Question 4
